\Large{Adversarial Training against Systematic Uncertainty}

### 2016-09-23 tex/paper.tex
save

The discovery of new particles or physical phenomena in high energy physics
experiments, like those at the Large Hadron Collider, requires the observation
of statistically significant deviations from the predictions of the Standard
Model, the current model of the known fundamental particles and their
interactions.  Typically, discovery requires satisfying the \textit{5-sigma
rule}, whereby the deviation must be at least five standard deviations from
predictions, i.e. having a p-value of $p \lesssim 3 \times 10^{-7}$. The
evaluation of the statistical significance must not only account for statistical
uncertainties, but also systematic uncertainties that represent our incomplete
knowledge of the theory of particle interactions or of the properties of the
experimental detection apparatus. Systematic uncertainties can alter the
expected rate of signal and background classes, e.g. their priors, as well as
the distributions of features. As such, the signal class true-positive rate and
the background class false-positive rate of a classifier can change with
variations in the systematic uncertainty thereby decreasing the sensitivity of
an analysis. In this paper, we present a new strategy for training a classifier,
based on adversarial training~\citep{goodfellow2014generative}, which aims to
build a classifier whose decision function is invariant under variations due to
systematic uncertainties.  In this way, uncertainties on predictions from
simulations will have a reduced impact when computing the significance of
potential deviations after application of the discriminant model and thereby
increase the sensitivity of the analysis to signals from potential new physics
sources. \glnote{Shall we be more explicit somewhere and put a formula
underlining the effect of having a pivotal classifier on the resulting
sensitivity? (versus a classifier whose decision function depends on the
nuisance.)}

@@

Since we do not have training tuples $(X_{z}(\omega),
X_{z^\prime}(\omega))$ (for the same unknown $\omega$), we propose instead to
solve the closely related problem of finding a predictive function $f$ such that
\begin{align}\label{eqn:criterion-measure}
    &P(\{ \omega | f(X_{z}(\omega) ; \theta_f) = y \} ) \nonumber \\
    &\indent = P( \{ \omega' | f(X_{z^\prime}(\omega') ; \theta_f) = y\})
\end{align}
for all $y \in {\cal Y}$. In words, we are looking for a predictive function $f$
which is a pivotal quantity \citep{degroot1986probability} with respect to the
nuisance parameter. That is, such that  the distribution of $f(X_z; \theta_f)$
is invariant with respect to the value $z$ of the nuisance. Note that a function
$f$ for which Eqn.~\ref{eqn:criterion-true} is true necessarily satisfies
Eqn.~\ref{eqn:criterion-measure}. In general, the converse is however not true,
since the sets of samples $\{ \omega | f(X_{z}(\omega); \theta_f) = y \}$ and $\{
\omega' | f(X_{z^\prime}(\omega'); \theta_f) = y \}$ do not need to be the same
for the equality to hold. In order to simplify notations, and as only
Eqn.~\ref{eqn:criterion-measure} is of direct interest in this work, we denote
from here on the pivotal quantity criterion as
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) | z ) = p(f(X ; \theta_f) | z^\prime )
\end{equation}
for all $z,z^\prime \in  {\cal Z}$.

@@

\begin{proof}
The optimal parameters $$\smash{\hat\theta}_r =
\arg \max_{\theta_r} E(\theta_f, \theta_r) = \arg \min_{\theta_r} {\cal L}_r(\theta_f,
\theta_r)$$ are such that $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z|f(X;\theta_f))$.
By assumption, $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$,
and therefore $p(z|f(X;\theta_f)) = p(z)$.
Using the Bayes' rule, we write
\begin{align*}
    p(f(X;\theta_f)|z) &= \frac{ p(z|f(X;\theta_f)) p(f(X;\theta_f)) } { p(z)}\\
                       &= p(f(X;\theta_f)),
\end{align*}
which holds for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$ and implies that $f$ is a pivotal quantity.
\end{proof}

@@

>>>>

The discovery of new particles or physical phenomena in high energy physics
experiments, like those at the Large Hadron Collider, requires the observation
of statistically significant deviations from the predictions of the Standard
Model, the current model of the known fundamental particles and their
interactions.  Typically, discovery requires satisfying the \textit{5-sigma
rule}, whereby the deviation must be at least five standard deviations from
predictions, i.e. having a p-value of $p \lesssim 3 \times 10^{-7}$. The
evaluation of the statistical significance must not only account for statistical
uncertainties \glnote{Shouldnt we define what are statistical uncertainties, so
that systematic uncertainties are in constrast better understood?}, but also
systematic uncertainties that represent our incomplete knowledge of the theory
of particle interactions or of the properties of the experimental detection
apparatus. Systematic uncertainties can alter the expected rate of signal and
background classes, e.g. their priors, as well as the distributions of features.
As such, the signal class true-positive rate and the background class
false-positive rate of a classifier can change with variations in the systematic
uncertainty thereby decreasing the sensitivity of an analysis. In this paper, we
present a new strategy for training a classifier, based on adversarial
training~\citep{goodfellow2014generative}, which aims to build a classifier
whose decision function is invariant under variations due to systematic
uncertainties.  In this way, uncertainties on predictions from simulations will
have a reduced impact when computing the significance of potential deviations
after application of the discriminant model and thereby increase the sensitivity
of the analysis to signals from potential new physics sources. \glnote{Shall we
be more explicit somewhere and put a formula underlining the effect of having a
pivotal classifier on the resulting sensitivity? (versus a classifier whose
decision function depends on the nuisance.)}

@@

Since we do not have training tuples $(X_{z}(\omega),
X_{z^\prime}(\omega))$ (for the same unknown $\omega$), we propose instead to
solve the closely related problem of finding a predictive function $f$ such that
\begin{align}\label{eqn:criterion-measure}
    &P(\{ \omega | f(X_{z}(\omega) ; \theta_f) = y \} ) \nonumber \\
    &\indent = P( \{ \omega' | f(X_{z^\prime}(\omega') ; \theta_f) = y\})
\end{align}
for all $y \in {\cal Y}$. In words, we are looking for a predictive function $f$
which is a pivotal quantity \citep{degroot1986probability} with respect to the
nuisance parameter. That is, such that  the distribution of $f(X_z; \theta_f)$
is invariant with respect to the value $z$ of the nuisance. Note that a function
$f$ for which Eqn.~\ref{eqn:criterion-true} is true necessarily satisfies
Eqn.~\ref{eqn:criterion-measure}. In general, the converse is however not true,
since the sets of samples $\{ \omega | f(X_{z}(\omega); \theta_f) = y \}$ and $\{
\omega' | f(X_{z^\prime}(\omega'); \theta_f) = y \}$ do not need to be the same
for the equality to hold. In order to simplify notations, and as only
Eqn.~\ref{eqn:criterion-measure} is of direct interest in this work, we denote
from here on the pivotal quantity criterion as
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) | z ) = p(f(X ; \theta_f) | z^\prime )
\end{equation}
for all $z,z^\prime \in  {\cal Z}$ and all values of $f(X ; \theta_f)$.

@@

\begin{proof}
Since we assume enough capacity for $r$, the optimal parameters $$\smash{\hat\theta}_r =
\arg \max_{\theta_r} E(\theta_f, \theta_r) = \arg \min_{\theta_r} {\cal L}_r(\theta_f,
\theta_r)$$ are such that $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z|f(X;\theta_f))$.
By assumption, $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$,
and therefore $p(z|f(X;\theta_f)) = p(z)$.
Using the Bayes' rule, we write
\begin{align*}
    p(f(X;\theta_f)|z) &= \frac{ p(z|f(X;\theta_f)) p(f(X;\theta_f)) } { p(z)}\\
                       &= p(f(X;\theta_f)),
\end{align*}
which holds for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$ and implies that $f$ is a pivotal quantity.
\end{proof}

@@

As an illustrating example, let us consider...
